### 2015-09-06 manuscript/content/abstract.tex
Revisions

\begin{abstract}
This paper proposes an application that collects sensor data from smartwatches and smartphones in order to predict blood alcohol content (BAC) or drunkenness in real-time, discreetly, and non-invasively through a machine learning approach. This system could prevent drunk driving or other dangers related to the consumption of alcohol by giving users a way to measure personal BAC without the use of intrusive breathalyzers or guess work.  Using data collected by a custom application from many volunteers, we trained a machine learning model that will work with an Android application to predict the user's BAC in real-time.
\end{abstract}

>>>>

\begin{abstract}
This paper proposes an application that collects sensor data from a smartwatch in order to predict drunkenness in real-time, discreetly, and non-invasively via a machine learning approach. This system could prevent drunk driving or other dangers related to the consumption of alcohol by giving users a way to determine personal intoxication level without the use of intrusive breathalyzers or guess work.  Using smartwatch data collected from several volunteers, we trained a machine learning model that may work with a smartphone application to predict the user's intoxication level in real-time.
\end{abstract}

### 2015-09-06 manuscript/content/conclusions.tex
Revisions

We took a look at the design of a general system that could be used for many applications, whether it be a BAC warning system, or a geographical mapping service that displays the smartwatch data of users on a chloropleth map. Part of this system was a general Android-based gateway that can be used to collect data from a variety of sensors connected to an Android smartphone.

After we collected some data, we spent some time to analyze it and look for interesting patterns that were immediately obvious. We found that skin temperature was a good indicator of drunkenness (in our controlled setting). We also discovered that excited heart rate looks to also be a indicator of intoxication level.

Following the overview analysis, we dove into training two regression and classification models. Achieving good performance as a regression problem was difficult. We found that the problem was much better tackled as a classification problem. This worked better because our classification models could ignore a good deal of the variance in the straight BAC predictions. In the end, we found SVM to perform the best on our data.

>>>>

We investigated the design of a general system that could be used for many applications, whether it be a BAC warning system, or a geographical mapping service that displays the smartwatch data of users on a chloropleth map. Part of this system was a general Android-based gateway that can be used to collect data from a variety of sensors connected to an Android smartphone.

We then designed an experiment to collect labeled data consistently from volunteers. After this, we analyzed the data to discover if there were any interesting patterns that were immediately obvious. We found that skin temperature was a good indicator of drunkenness (in our controlled setting). We also discovered that excited heart rate looks to also be a good indicator of intoxication level.

Following the overview analysis, we dove into training some regression and classification models. Achieving good performance as a regression problem was difficult. We found that the problem was much better tackled as a classification problem. This worked better because our classification models could ignore a good deal of the variance in the straight BAC predictions. In the end, we found SVM to perform the best on our data.

### 2015-09-06 manuscript/content/eval.tex
Revisions

In this section we will take a look at the performance of several different models using the aforementioned features. We will start off with some regression modeling and then move on to some classification modeling. Because of our small number of volunteers, we are unlikely to see great results; though we can at least get an idea of the performance of the models and features we use.

@@

The first model we will consider a classic and a good starting point: the basic linear model. For these predictions, we split the data 80\% for training and 20\% for testing. The following performance values are the values calculated on the unseen test data. Using only the heart rate data, we found the model acheived extremely low performance, with an R$^2$ of 0.077 and root mean squared error (RMSE) of 0.036. Ideally, we want an R$^2$ value of 1.0 and an RMSE of 0.0. Adding in the skin temperature data boosted the R$^2$ up to 0.268 with 0.032 RMSE. Better, but still bad. Adding in the accelerometer data the values were 0.276 and 0.032, for R$^2$ and RMSE respectively. Even using all our preselected features, the linear model is just not a good fit.

So how does a more flexible model perform? We next trained a neural network model on the data. Training a neural network model is very expensive in terms of time, so we reduced our dataset to 8\% of the original. Of the remaining amount we used 70\% for training the model and 30\% for testing. The neural network itself had two hidden layers of five nodes each. This structure was determined experimentally while optimizing for training time and performance. In the input layer we used the heart rate, skin temperature, and accelerometer features. Figure \ref{fig:nn_all} plots the BAC predictions in of the neural network model on top of the actual values from the test set. The real values are jittered to give a sense of the density of the data, the prediction values in blue are not. Overall, the model achieved a much better R$^2$ value of 0.525, with RMSE of 0.026, than the linear regression model.

@@

This problem might be better handled as a classification task. To do this we will first have to use a threshold to classify the existing BAC labels as \textit{DRUNK} or \textit{SOBER}. We want to warn a person if they are close to reaching the legal limit of 0.08 BAC. A good time to warn a user is at about 0.065 BAC. This splits the classes into 64\% is drunk, and 36\% sober.

The common logistic regression model will be the first to try for our classification task. To train the model, we split the data set by 80\% for training and 20\% for testing as we did for the linear regression model. We will also go ahead and use the full combination of heart, temperature, and accelerometer features. A logistic regression model outputs values from 0.0 to 1.0, so we need to determine where to best split this output into each class. In Figure \ref{fig:log_pred_density}, we show a plot of the predictions of the model on the test data, using the actual labels to distinguish the output. In this plot, we see that the best threshold value for the logistic model predictions is around 0.32; above that we classify \textit{SOBER} and below that \textit{DRUNK}. Using this split, we achieved a precision score of 0.852, recall of 0.730, and F1-score of 0.786.

The final model we will consider is the SVM using the Gaussian Radial Basis Function (RBF) kernel. This time we first sampled 50\% of the data and then split this into 80\% for training and 20\% for testing. The initial reduction is performed because the SVM is very time-consuming to train, a bit faster than the ANN. On our 0.065-BAC-thresholded data, we found the SVM acheived great performance with a precision of 0.887, recall of 0.931, and overall F1-score of 0.909 on the test data. Ideally, we would not want to warn our users that they are drunk when they are not actually drunk, so we want to try and optimize the model to be as precise as possible. By modifying the error weighting to train against false positive errors, the SVM model achieved a precision of 0.974, with recall of 0.732, and F1-score of 0.836. Our recall dropped for a higher precision, which is a well-known tradeoff for this kind of tuning, but this is fine. We can set a threshold on our smartphone warning system that if at least a recall fraction of the samples from our smartwatch are classified as \textit{DRUNK}, then there is a precision chance that the user is actually at 0.065 BAC.

>>>>

Overall, our data set contains 233,538 samples that were collected from five volunteers. Each sample was collected every three seconds from a Microsoft Band smartwatch by our Android data collection system. In this section, we evaluate the performance of a few machine learning models on our dataset. We use some standard performance measures for our evaluation: precision, recall, and F1-score, for the classification models; RMSE and R$^2$ for the regression models. All reported performance values are determined via 5-fold cross-validation.

We want to warn a person if they are close to reaching the legal limit of 0.08 BAC. A good time to warn a user is at about 0.065 BAC. Using this threshold, the classes are split into 64\% is \textit{DRUNK}, and 36\% \textit{SOBER}.

To get a baseline, we first trained a logistic regression model. This model outputs values from 0.0 to 1.0, so we need to determine where to best split this output into each class. In Figure \ref{fig:log_pred_density}, we show a plot of the predictions of the model on the test data, using the actual labels to distinguish the output. In this plot, we see that the best threshold value for the logistic model predictions is around 0.32; above that we classify \textit{SOBER} and below that \textit{DRUNK}. Using this, we achieved a precision score of 0.852, recall of 0.730, and F1-score of 0.786.

Moving forward, we trained a SVM model using the Gaussian Radial Basis Function (RBF) kernel. For time constraints, the dataset was reduced by half using a uniform subsample. Even so, we found the SVM was able to acheive great performance with a precision of 0.887, recall of 0.931, and overall F1-score of 0.909. Ideally, we would not want to warn our users that they are drunk when they are not actually drunk, so we want to try and optimize the model to be as precise as possible. By modifying the error weighting to train against false positive errors, the SVM model achieved a precision of 0.974, with recall of 0.732, and F1-score of 0.836. Our recall dropped for a higher precision, which is a well-known tradeoff for this kind of tuning, but this is fine. We can set a threshold on our smartphone warning system that if at least a recall fraction of the samples from our smartwatch are classified as \textit{DRUNK}, then there is a precision chance that the user is actually at 0.065 BAC.

@@

So how does this do as a regression problem? We first considered the most basic: a linear regression model. This did not perform well at all. We next trained a neural network (ANN) model on the data. This was even slower than the SVM which forced us to subsample to 8\% of the original data. The best performing ANN had a structure with two hidden layers of five nodes each. Figure \ref{fig:nn_all} plots the BAC predictions in of the neural network model on top of the actual values from the test set. The real values are jittered to give a sense of the density of the test data, the prediction values in blue are not. This model acheived a R$^2$ value of 0.525, with RMSE of 0.026. Much better than the linear regression model, but not nearly as good as the classification models.

@@

### 2015-09-06 manuscript/content/method.tex
Revisions

All work will be performed in R version 3.2.1 using the additional 'neuralnet' and 'kernlab' packages. Any necessary information will be provided in the evaluation section. To evaulate the models will use some standard performance measures: precision, recall, and F1-score, for the classification models; RMSE and R$^2$ for the regression models.

>>>>

All work will be performed in R version 3.2.1 using the additional 'neuralnet' and 'kernlab' packages. Any necessary additional information will be provided in the evaluation section.

### 2015-09-06 manuscript/content/system.tex
Revisions

Before we get into the specifics of looking into the data and the prediction of the intoxication level, let's take a quick look at the underlying infrastructure of our collection system. A huge end-goal of ours is to have a public system where participating users' phones transmit sensor information to a central database through a SOAP-based web service. The service stores this information in the database and another data processing system. Other users who are not contributing information can connect to our system to access other related services, like a chloropleth map of the sensor data; BAC prediction data in our case, but the system would not limited to it. The particular protocols and setup used should be designed to protect the privacy of the sensor data contributors.

>>>>

Before we get into the specifics of looking into the data and the prediction of the intoxication level, let's take a quick look at the underlying infrastructure of our collection system. A huge end-goal of ours is to have a public system where participating users' phones transmit sensor information to a central database through a REST-based web service. The service stores this information in the database and another data processing system. Other users who are not contributing information can connect to our system to access other related services, like a chloropleth map of the sensor data; BAC prediction data in our case, but the system would not limited to it. The particular protocols and setup used should be designed to protect the privacy of the sensor data contributors.
